`ggmatplot`: An R package for data visualization on wide-format data

### 2022-03-02 Paper/paper.Rmd
rewrote statement of need and added example data

@@

The layered grammar of graphics [@Wickham2010-kt], implemented as the `ggplot2` package [@Wickham2016] in the statistical language R [@rstats], is a powerful tool to create versatile statistical graphics. This graphical system, however, requires input data to be organised in a manner that a data column is mapped to an aesthetic element (e.g. x-coordinate, y-coordinate, color, size), which create friction in constructing plots with an aesthetic element that span multiple columns in the data by requiring users to re-organise the data.

The `ggmatplot`, built upon `ggplot2`, is an R-package that allows quick plotting across the columns of matrics or data with the result returned as a `ggplot` object. The package is inspired by the function `matplot()` in the core R `graphics` system, thus `ggmatplot` can be considered as a `ggplot` version of `matplot` with the benefits of customising the plots as any other `ggplot` objects via `ggplot2` functions.

@@

Data can be tidied in a rectangular form where each row represents an observational unit, each column represents a variable, and each cell represents a value [@Wickham2014-gy]. What constitutes a variable, hence a column, in a tidy data are dependent upon interpretation.

---

The increased use of a tidy approach in structuring datasets results in wrangling datasets into wide format: with variables as columns and observations as rows(https://vita.had.co.nz/papers/tidy-data.html). While this introduces a more meaningful mapping of the data into its structure, the grammar of graphics requires wide format data to be wrangled back into long format for visualizing. This can be quite cumbersome when creating simple plots, and converting a tidy data frame back to a long format with multiple variables stacked upon each other can also be considered counter-intuitive.

Therefore, the motivation for `ggmatplot` is to provide a solution that allows `ggplot2` to handle wide format data. Although `ggmatplot` doesn't provide the same flexibility as `ggplot2`, it can be used as a workaround for having to wrangle wide format data into long format and creating simple plots using `ggplot2`.

@@

>>>>

```{r setup, include = FALSE}
library(tidyverse)
library(knitr)
library(kableExtra)
knitr::opts_chunk$set(message = FALSE,
                      warning = FALSE)

# data

wide_df <- tribble(~restaurant, ~person, ~food_rating, ~service_rating, ~ambience_rating, ~overall_rating,
        "R1", "P1", 4, 3, 4, 4,
        "R1", "P2", 4, 5, 4, 4,
        "R1", "P3", 3, 4, 5, 3, 
        "R2", "P1", 2, 4, 4, 3, 
        "R2", "P2", 3, 4, 4, 3)

long_df <- pivot_longer(wide_df, contains("rating"), 
                        names_to = "rating_type",
                        names_pattern = "(.+)_rating",
                        values_to = "rating")

abund_df <- tribble(~Site, ~`Soil dry mass`, ~`Moss`, ~Alopcune, ~Arctlute, ~Pardpull, ~Trocterr, ~Zoraspin,
                    1, 2.3321, 3.0445, 10, 0, 45, 57, 4,
                    2, 3.0493, 1.0986, 2, 0, 37, 65, 9,
                    3, 2.5572, 2.3979, 20, 0, 45, 66, 1)

table_styling <- function(df, ...) {
  kable(df, ..., booktabs = TRUE, linesep = "") %>% 
    kable_classic()
}

data("spider", package = "mvabund")
```

@@

The layered grammar of graphics [@Wickham2010-kt], implemented as the `ggplot2` package [@Wickham2016] in the statistical language R [@rstats], is a powerful and popular tool to create versatile statistical graphics. This graphical system, however, requires input data to be organised in a manner that a data column is mapped to an aesthetic element (e.g. x-coordinate, y-coordinate, color, size), which create friction in constructing plots with an aesthetic element that span multiple columns in the original data by requiring users to re-organise the data.

The `ggmatplot`, built upon `ggplot2`, is an R-package that allows quick plotting across the columns of matrices or data with the result returned as a `ggplot` object. The package is inspired by the function `matplot()` in the core R `graphics` system, thus `ggmatplot` can be considered as a `ggplot` version of `matplot` with the benefits of customising the plots as any other `ggplot` objects via `ggplot2` functions.

@@

Input data to construct plots with `ggplot2` require data to be organised in a manner that maps data columns to aesthetic elements. This required form is, however, not consistent with some common data formats, specifically data where the column names contain information to be mapped into the plot or the variables to be mapped into the same aesthetic elements. Consequently, plotting with `ggplot2` interrupts the workflow of a user that is trying to quickly explore these types of data. The `ggmatplot` R-package provides a solution to this common friction to produce plots with `ggplot2`.

@@

Data can be tidied in a rectangular form where each row represents an observational unit, each column represents a variable, and each cell represents a value [@Wickham2014-gy]. In some cases, what constitutes a variable (or observational unit), hence a column (or row), in a tidy data can be dependent upon interpretation or downstream interest (e.g. Tables \ref{tab:tab1} and \ref{tab:tab2} can be both considered as tidy data), but a clear violation of tidy data principles is when the column names contain data values, e.g. Table \ref{tab:tab3} contain the name of the species across a number of column names.

```{r tab1, echo = FALSE}
table_styling(wide_df, col.names = c("Restaurant", "Person", "Food",
                                     "Service", "Ambience", "Overall"),
              caption = 'Restaurant rating data in "tidy" form') %>% 
  add_header_above(c(" ", " ", "Rating" = 4))
```

```{r tab2, echo = FALSE}
table_styling(long_df, col.names = c("Restauant", "Person", "Rating type", "Rating"),
              caption = 'Restaurant rating data in "molten" form')
```

```{r tab3, echo = FALSE}
table_styling(abund_df,
              caption = 'Spider abundance data with environmental covariates.') %>% 
  add_header_above(c(" ", "Environment covariates" = 2, "Species abundance" = 5))
```
